---
title: "Introduction to R - Template script"
output: html_document
date: "2024-11-06"
---

# Setting up

[You can write your notes as normal text like this, outside the code chunks]

## Load packages

```{r load-packages}


```

## Load and preview data

```{r load-data}


```

# Select variables

```{r select-examples}


```

**EXERCISE 1:** Create a new data frame called `selected_vars` which includes the following six variables: ID, learnTime, time, item, acc, RT

```{r select-exercise}


```

# Rename variables

Here's a list of alternative column names: "ppt_ID", "read_group", "exp_task", "counterbalanced_order", "encode_time", "test_time", "word", "correct", "rt"

```{r rename-examples}

```

**EXERCISE 2:** Create a new dataframe called `renamed_vars`, by renaming your `selected_vars` dataframe from Exercise 1. Edit the column names for the following variables:

-   ID >> ppt_ID
-   learnTime >> learn_time
-   time >> test_time

```{r rename-exercise}


```

# Filter variables

```{r filter-examples}


```

**EXERCISE 3:** What do you think the following lines of code might do to the data? Talk through your predictions, then run the code and inspect the dataframes to see if you are correct. You can edit the code comments (after the #) to make notes.  

```{r filter-exercise}
# Example 1
filter_dat2 <- filter(raw_dat, RT > 500)

# Example 2
filter_dat3 <- filter(raw_dat, RT <= 2000)

# Example 3
filter_dat4 <- filter(raw_dat, !is.na(RT))

# Example 4
filter_dat4 <- filter(raw_dat, item == "agouti")

# Example 5
filter_dat5 <- filter(raw_dat, time == 1 & RT <= 2000 )

# Example 6
filter_dat6 <- filter(raw_dat, item == "banyan" | item == "caracal")
```

# Combining processing steps using the pipe

```{r pipe-example}



```

# Calculate participant averages 

```{r summarise-examples}



```

**EXERCISE 4**: Calculate group summary statistics on mean RT scores at each test point. You can use the previous code as a starting point. Work out the mean and standard deviation, then if you have time you can also add the minimum and maximum response times.

```{r summarise-exercise}


```

# Create new variables

```{r new-var-examples}



```


# Putting it all together

**EXERCISE 5:** Create a dataset called `sleep_wake` by chaining the following steps together:

1.  Take the raw data
2.  Include only ID, learnTime, time, acc, and RT (renaming if you choose)
3.  Subset the data to include the first two test times only
4.  Create a log-transformed RT variable (log_RT), by using the log() function on RT
5.  Group by ID, learnTime and time to calculate average accuracies, raw RTs and log-RTs (i.e., for each participant for each condition)

```{r final-exercise-ppt}


```

Once you have done this, print the whole group summaries by learnTime and time. How does performance change between time 1 and 2 when participants learn in the PM versus the AM?

```{r final-exercise-group}


```

## Save data

```{r save-data}


```

